---
title: "Simplified_biclust_with_likelyhood_assignmend_walkthrough"
output: html_document
date: "2023-aug-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = FALSE)
library(tidyverse)
library(aricode)  # To calculate rand index
library(ggplot2)
library(ggalluvial)
library(reshape2)
library(here)  # To work with paths

path_root <- here::here()
execution_path <- here::here("simplified_biclust_likelihood")
output_path <- execution_path
function_folder <- here::here("functions")
all_function_files <- list.files(function_folder, recursive=T, full.names=T)
for(current_file in all_function_files){
    print(paste("Loading", current_file))
    source(current_file)
  }


#plot cluster evolution
#  plot_cluster_history <- function(cell_cluster_history){
# 
#   d <- cell_cluster_history
#   d <- d[ , colSums(is.na(d))==0]  # Remove NA data
#   d <- melt(d, id.vars="Cell ID")
#   colnames(d) <- c("cell", "iteration", "cluster")
#   d['cluster'] <- as.factor(d[, 'cluster'])
# 
#   rand_ind <- RI(cell_cluster_history[,2],cell_cluster_history[,ncol(cell_cluster_history)])
# 
#   # Plotting it
#   # Slow. But keeps track of individual cells
#   # ggplot(d, aes(x = iteration, stratum = cluster, alluvium = cell, fill = cluster, label = cluster)) +
#   #   scale_fill_brewer(type = "qual", palette = "Set2") +
#   #   geom_flow(stat = "alluvium", lode.guidance = "rightleft", color = "darkgray") +
#   #   geom_stratum() +
#   #   theme(legend.position = "bottom") +
#   #   ggtitle("Cluster allocation for each iteration")
# 
#   # Doesn't keep track of individual cells
#   ggplot(d, aes(x = iteration, stratum = cluster, alluvium = cell, fill = cluster, label = cluster)) +
#     scale_fill_brewer(type = "qual", palette = "Set2") +
#     geom_flow() +
#     geom_stratum() +
#     ylab("Cells") +
#     xlab("Iteration") +
#     labs(fill="Cluster") +
#     theme(legend.position = "bottom") +
#     ggtitle(paste0("Log of cluster allocation\nRand index of true vs final: ",round(rand_ind,2)))
# 
# }
# 

# Plot histograms of r2
r2plot <- function(iteration, r2, prev_cell_clust, n_cell_clusters = 2){
  if(T){
    par(mfrow=c(length(unique(prev_cell_clust)),n_cell_clusters))
    for(i_cells_from_cell_cluster in 1:length(unique(prev_cell_clust))){
      for(i_fits_into_cell_cluster in 1:length(unique(prev_cell_clust))){
        print(paste(i_cells_from_cell_cluster, i_fits_into_cell_cluster))
        ind_for_cell_cluster = which(rep(1:n_cell_clusters, n_target_gene_clusters)==i_fits_into_cell_cluster)
        hist(r2[ prev_cell_clust==i_cells_from_cell_cluster, ind_for_cell_cluster], breaks=100, main=paste("Cells from cell cluster", i_cells_from_cell_cluster, "\nfits into cell cluster", i_fits_into_cell_cluster, "with r2:"))
      }
    }
    mtext(paste("Iteration", iteration), side = 3, line = -1, outer = TRUE)
  }
}

set.seed(1234)

```

## Simplified biclust demo

The idea of this script is to walk through one iteration of our simplified version of biclust and find out either why the concept is flawed or if we are missing something that should be obvious.

The main idea is to consider biclust as a modular clustering algorithm that in general should be able to take in any kind of linear model-like method and use it for clustering. The core idea is that we have something that builds a linearisch model that we can use to derive r2 for cells in the cluster, and predicted r2 values for cells in other clusters (those that were not used for fitting that particular models), so we can use these r2/predicted r2 to reassign cells to clusters and re-fit models.

For this ultra-simple example, we will try this concept using the simplest thinkable method: a linear model of one dependent and one independent variable.

## Simulating some data
First, we build some gene expression for some regulator genes.

```{r regulators}
num_cells <- 100
regulator_expression <- rnorm(num_cells, mean = 1, sd = 0.1)
hist(regulator_expression)

```

Now using these regulator gene expressions we will build some target gene expressions. We want the target gene expressions to be clearly clustered for this example.
This will correspond to them being built from different regulatory models. So, first we define how many clusters we want, and define our regulatory model for each cell cluster.

```{r models}

n_cell_clusters    <- 2

# Sample intercepts
intercepts <- rnorm(n_cell_clusters, mean = 10, sd = 5) * 0 #can just set to zero without loss of generality as data will be 
# Sample slopes
betas      <- rnorm(n_cell_clusters, mean = 0,  sd =  5)
#target-gene variance per target gene cluster
sigmas     <- rnorm(n_cell_clusters, mean = 1,  sd = 0.1)


# Assign half of the cells to be in either cluster
true_cell_clust <- c(rep(1,num_cells/2), rep(2,num_cells/2))

```

Next we use these parameters to generate target gene expression.

```{r target expression ,fig.show="hold", out.width="50%"}

# Build cell expression from linear model
# | intercept, beta | x | r11, r12 |
#                       | r21, r22 |
#                       |    ...   |
# For every cell cluster, 1x2 * 2x50
cell_cluster_expression <- sapply(1:n_cell_clusters, function(cellCluster) 
  (
    c(intercepts[cellCluster], betas[cellCluster]) %*%
                                    t(cbind(rep(1, num_cells/2), regulator_expression[which(true_cell_clust == cellCluster)])) 
    ) + #add noise
    rnorm(num_cells/2, 0, sigmas[cellCluster])
)

target_expression <- c(cell_cluster_expression[,1], cell_cluster_expression[,2])
dat <- cbind(target_expression, regulator_expression)
ind_targetgenes <- c(1)
ind_reggenes <- c(2) 

hist(target_expression, breaks = 100)

p <- ggplot(as.data.frame(cbind(dat[,], true_cell_clust)), aes(x = regulator_expression, y = target_expression, color=factor(true_cell_clust)))

p + geom_point()

```
The end goal of the real biclust is that it can use another clustering method to find initial cluster labels, and then hopefully it could improve upon this using the additional information from scregclust

For this simple example, our idea is to put some observations into the wrong cluster, and see if the biclust-algorithm can put them back to the true cluster. We set up cluster labels with mostly true labels, but a proportion of the labels permuted.

```{r}

disturbed_initial_cell_clust <- true_cell_clust
disturbed_fraction <- 0.20
for(i_cluster in 1:n_cell_clusters){
  indexes_of_cluster <- which(true_cell_clust == i_cluster)
  some_of_those_indexes <- sample(indexes_of_cluster, size=as.integer(length(indexes_of_cluster)*disturbed_fraction), replace = F)
  disturbed_initial_cell_clust[some_of_those_indexes] <-
    sample(c(1:n_cell_clusters)[-i_cluster],
           size=length(some_of_those_indexes), replace=T)
}


cell_cluster_history <- cbind(true_cell_clust, disturbed_initial_cell_clust)
colnames(cell_cluster_history) <- c("True allocation", "Disturbed allocation")

```

## Biclust, first iteration
Now it's time to actually do something resembling biclust.

The real thing would obviously work for many iterations, but for now lets just walk through one iteration.

First we set up and pre-allocate some variables that we'll use later.

```{r}
i_main <- 1  # Main iteration variable, just set it to one for now.
max_iter <- 1  # This one would usually cap number of iterations, set to 1 for now


# Find initial cluster labels
initial_clustering <- disturbed_initial_cell_clust
n_target_gene_clusters <- 1  # We are not clustering target genes for now and we only have one target gene


# Set up some variables
n_cell_clusters = length(unique(initial_clustering))
n_target_genes = 1
n_regulator_genes = 1


# Set up cluster history
initial_column_padding <- ncol(as.matrix(initial_clustering)) + 1  # +1 Because we have an index column that is not an index column it's an ID column
cell_cluster_history <- data.frame(matrix(NA, nrow = nrow(as.matrix(initial_clustering)), ncol = max_iter + initial_column_padding))
colnames(cell_cluster_history) <- c("Cell ID", 'Initial clustering', paste0("Iteration ", 1:max_iter))
cell_cluster_history[, 'Cell ID'] <-1:length(initial_clustering)  # Set cell names
cell_cluster_history[, 'Initial clustering'] <- initial_clustering


# Pre-allocate all r2 matrices for later analysis if feasible
like_all <- vector("list", length = max_iter)

# Set flag for breaking out of the loop.
stop_iterating_flag = T

# Set the current cell clustering
current_cell_cluster_allocation <- initial_clustering

```

First step is to, for each cell cluster separately, fit a standard linear model.

As an inbetween, we can check if doing this on the true cell clusters retrieves the intended models:
```{r}

models <- vector("list", length = n_cell_clusters)
dat <- cbind(dat, NA)  # Add columns so we can have colors for clusters in the plot
colnames(dat)[3] <- 'cluster'  # Name that column
for(cell_cluster in 1:n_cell_clusters){
  current_rows <- which(true_cell_clust == cell_cluster)
  dat[current_rows, 'cluster'] <- cell_cluster  # Again, so we can plot the colors for each cluster
  models[[cell_cluster]] <- lm(dat[current_rows,'target_expression'] ~ dat[current_rows,'regulator_expression'])
}

p <- ggplot(as.data.frame(dat[,]), aes(x = regulator_expression, y = target_expression, color=factor(cluster))) +
  geom_point() +
  geom_abline(intercept = coef(models[[1]])[1], slope = coef(models[[1]])[2], color = "red") +
  geom_abline(intercept = coef(models[[2]])[1], slope = coef(models[[2]])[2], color = "cyan") + geom_abline(intercept = intercepts[1], slope = betas[1], color = "pink") +
  geom_abline(intercept = intercepts[2], slope = betas[2], color = "blue") +
  labs(title = "Scatterplot with Regression Lines and true model lines")

# Reset changes used for this plot
dat <- dat[,1:2]

#plot output
p + labs(color = "True cluster allocation")


```


```{r  fig.show="hold", out.width="50%"}
# Fit model to each cell cluster
models <- vector("list", length = n_cell_clusters)
dat <- cbind(dat, NA)  # Add columns so we can have colors for clusters in the plot
colnames(dat)[3] <- 'cluster'  # Name that column
for(cell_cluster in 1:n_cell_clusters){
  current_rows <- which(current_cell_cluster_allocation == cell_cluster)
  dat[current_rows, 'cluster'] <- cell_cluster  # Again, so we can plot the colors for each cluster
  models[[cell_cluster]] <- lm(dat[current_rows,'target_expression'] ~ dat[current_rows,'regulator_expression'])
}

p <- ggplot(as.data.frame(dat[,]), aes(x = regulator_expression, y = target_expression, color=factor(cluster))) +
  geom_point() +
  geom_abline(intercept = coef(models[[1]])[1], slope = coef(models[[1]])[2], color = "red") +
  geom_abline(intercept = coef(models[[2]])[1], slope = coef(models[[2]])[2], color = "cyan") +
  labs(title = "Scatterplot with Regression Lines")

# Reset changes used for this plot
dat <- dat[,1:2]

#plot output
p + labs(color = "Starting cell cluster allocation")

```

Now that we have fitted some models, it's time to re-allocate cells to appropriate clusters.

To do this, we calculate the probability of each cell to follow from each linear model. This is an extremely simplified version of what we will later look into.

Since there is only one regulator gene, we will skip the L1 penalty for adding coefficients that we would use in a larger model, and instead only look at the negative log-likelihood.


```{r}

# For all cells, calculate the likelihood of coming from the model corresponding to each 
like <- matrix(0, nrow = nrow(dat), ncol = n_cell_clusters)

penalization_LAMBDA <-  0.5

# Calculate the residual target gene variance for each gene and cluster
# (so just one gene).
TARGET_GENE_RESIDUAL_var <-  matrix(0, nrow = n_target_genes, ncol = n_cell_clusters)
# dat is dat <- cbind(target_expression, regulator_expression), e.g. a 2x100, with e.g. the first 50 rows being true cell cluster 1
# 100x2 * 2x1 

  for(cell_cluster in 1:n_cell_clusters){
    #
    intercept_and_regulatorgenes <- as.matrix(cbind(rep(1,nrow(dat)),dat[,ind_reggenes]))  # 100x2
    current_rows <- which(current_cell_cluster_allocation == cell_cluster)
    
    residuals_temp <- dat[current_rows,ind_targetgenes] - (intercept_and_regulatorgenes[current_rows,] %*% models[[cell_cluster]]$coefficients)
    TARGET_GENE_RESIDUAL_var[cell_cluster]  <- var(residuals_temp)
  }
  
  # Now to actually calculate predicted or 'predicted' r2
  for(cell in 1:nrow(dat)){
    for(cell_cluster in 1:n_cell_clusters){
      # Bug fix hack: remove NA coefficients
      if(any(is.na(models[[cell_cluster]]$coefficients))){
        NA_coeffs <-  unname(which(is.na(models[[cell_cluster]]$coefficients)))
        S_ERR <- (dat[cell,ind_targetgenes] - as.vector(c(1,dat[cell,c(-1, -NA_coeffs)])) %*% models[[cell_cluster]]$coefficients[-NA_coeffs])^2
      }
      
      SQUARED_ERROR <- (dat[cell,ind_targetgenes] - as.vector(c(1,dat[cell,ind_reggenes])) %*% models[[cell_cluster]]$coefficients)^2
      
      penalization <- penalization_LAMBDA * models[[cell_cluster]]$coefficients[2]/TARGET_GENE_RESIDUAL_var[cell_cluster]
      
      # like[cell,cell_cluster] <- SQUARED_ERROR / 2 / TARGET_GENE_RESIDUAL_var[cell_cluster] - penalization #negative penalty as higher likelihood is better
      
      like[cell,cell_cluster] <- -log(TARGET_GENE_RESIDUAL_var[cell_cluster])/2 - 
                                  1/2/TARGET_GENE_RESIDUAL_var[cell_cluster]* SQUARED_ERROR + #here we are optimizing the penalized NEGATIVE likelyhood, so penalty is positive
                                  penalization 
      
    }
    
  }

  like_all[[i_main]] <- like

  r2plot(iteration = i_main,
       r2 = like,
       prev_cell_clust = true_cell_clust)

  p <- ggplot(data = as.tibble(cbind(like, true_cell_clust)),
       aes(x = V1, y = V2, color = as.factor(true_cell_clust))
      ) +
      geom_point() +
        geom_abline(intercept = 0, slope = 1)+
        labs(x = "Log-likelihood for fitting into cluster 1", y = "Log-likelihood for fitting into cluster 2")
  
  p + labs(color = "True cell cluster")
  

```


And now that we have our like data, it's time to re-assign cluster labels. This is straightforward, we just select the r2 column with the lowest r2, indicating best fit.
Usually this would raise the concern of over-fitting, since a cell was used to fit the model for the cluster it was in the start of the iteration. If that was a big problem, cells would tend to be assigned to the same cluster as they started in. As you can see, that's not really a problem.

```{r}

# Update cluster allocations
updated_cell_clust <-  sapply(1:nrow(like), function(row) which.min(like[row,]))

# Update data in cell_cluster_history
cell_cluster_history[, i_main + initial_column_padding] <- updated_cell_clust
RI(true_cell_clust,updated_cell_clust)

```


```{r}

cell_cluster_history_plotting <-cbind('Cell ID' = cell_cluster_history[,1],
                             true_cell_clust,
                             cell_cluster_history[,c(2,3)])

plot_cluster_history(cell_cluster_history = cell_cluster_history_plotting)

```


So in this super simple version the cells find their way back. 


